Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-05-06 Manuscript/manuscript.Rmd
manuscript edits

Current methods used to predict TAD boundaries rely on the use of classification algorithms to draw associations between the spatial relationship of functional genomic elements and regions that include TAD boundaries. In doing so, many methods ignore key characteristics of domain data, such as the impact of heavily imbalanced number of boundaries vs. nonboundary regions. This is because, when constructing the data for modeling, a common methodology used to determine the class of the response variable is to bin the linear genome and then create a method used to determine if each bin "includes" a TAD boundary or not. Given the sparcity of TADs throughout the genome though, this will inevitably lead to severly imbalanced classes.

In a classification setting, when one is attempting to discriminate between regions with (minority class) and without (majority class) TAD boundaries, the class imbalance issue will contribute to small imbalance ratios. The imbalance ratio is defined as the ratio of minority to majority classes [@Orriols-Puig2008]. Such heavily imbalanced datasets are known to adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class; @chen2013influence; @jeni2013facing]. Consequently, careful consideration must be taken when deciding which performance metric(s) to use for evaluation of predictive models, especially in the case of imbalanced data.

Commonly used measures of model performance for binary classifiers include threshold-free metrics like the area under the receiver operating curve (AUROC), and other threshold-specific metrics like accuracy, sensitivity, and specificity. The ROC curve is created by plotting the true positive rate (TPR) against the false positive rate (FPR) at various threshold settings. The subsequent AUROC, measures how well the model is capable of distinguishing between the two classes [@park2004receiver]. Accuracy is defined as the ratio of correctly identified cases (either from the positive or negative class) over the total number of cases. It measures the overall proportion of correctly classified instances. Sensitivity, also referred to as recall or TPR, is the ratio of true positives over the total number of positive classes. Specificity, on the other hand, also referred to as TNR, is the ratio of true negatives over the total number of negative classes. Each measures the proportion of correctly identified positive and negative classes respectively [@yerushalmy1947statistical]. Problems arise when using such measures in the event of imbalanced classes. For AUROC, when the prevalence of an event is low--that is, a defined genomic region truly containing a TAD boundary--mean probabilities are biased towards the majority class. This results in the area of interest under an ROC curve to be compressed toward a small corner in the ROC space [@prati2011survey; @wei2013role]. While not directly affecting the value of the AUROC, this can have severe consequences on the interpretability of the model's performance. A better threshold-free measure to use when dealing with imbalanced data is the area under the precision-recall curve (AUPRC).  The interpretation of the AUPRC of a model is similar to that of the AUROC. However, where as the baseline model performance according to the AUROC is given by 0.50, the baseline model performance according to the AUPRC is given by the ratio of true positives over the total number of cases. This enables a more realistic benchmark to evaluate a model on as this ratio will be lower than 0.5 for a test data set with imbalanced classes. Likewise, threshold-specific metrics like accuracy and specificity are also problematic when evaluating model performance on imbalance data due to being highly influenced by the correct classification of the majority class. A better measure to consider is a composite metric involving all quadrants of the confusion matirix, such as Matthew's Correlation Coefficient (MCC) [@jeni2013facing; @chicco2017ten]. The MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications [@brodersen2010balanced]. A value of positive 1 represents a perfect prediction, while a value of negative 1 represents total disagreement between prediction and observation. A value of 0 indicates no better than random prediction.

<!--
precision and recall together because they soften the influence of a model's ability to correctly classify majority classes and yield more dependable and interpretable results. Therefore, measures like the F1-Score and Matthew's Correlation Coefficient (MCC) offer better interpretation, especially in the instance of imbalanced classes [@jeni2013facing]. The F1-Score can be interpreted as a weighted average of the precision and recall values, while the MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications [@brodersen2010balanced]. Each metric uses all 4 quadrants of the confusion matrix, thus providing more balanced measures in the case of imbalanced data [@chicco2017ten].
-->

<!--
, and balanced accuracy offer better interpretation, especially in the instance of imbalanced classes [@jeni2013facing]. The F1-Score can be interpreted as a weighted average of the precision and recall values, balanced accuracy measures the average accuracy obtained on either class, and MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications [@brodersen2010balanced]. Each metric uses all 4 quadrants of the confusion matrix, thus providing more balanced measures in the case of imbalanced data [@chicco2017ten].
-->

<!--
Certain composite threshold-specific metrics such as balanced accuracy and Youden's Index should be preferred because they soften the influence of a model's ability to correctly classify majority classes and yield more dependable and interpretable results [@jeni2013facing]. Here, balanced accuracy measures the average accuracy obtained on either class [@brodersen2010balanced]. Informedness, also referred to as Youden's Index, measures the probability of the model making an informed decision and corresponds to giving equal weight per class, rather than to each instance like Accuracy [@youden1950index;@powers2011evaluation].
-->

<!--
Better measures to consider are composite metrics involving both precision and recall together. Therefore, measures like the F1-Score, Youden's Index (YI), and Matthew's Correlation Coefficient (MCC) offer better interpretation, especially in the instance of imbalanced classes [@jeni2013facing]. The F1-Score can be interpreted as a weighted average of the precision and recall values, while Youden's index is interpreted as the arithmetic mean between sensitivity (recall) and specificity. MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications. Each metric uses all 4 quadrants of the confusion matrix, thus providing more balanced measures in the case of imbalanced data [@chicco2017ten].
-->

In addition to employing suitable metrics for evaluating models validated on imbalanced data, there also exists several algorithmic or data level solutions. Algorithmic level solutions involve modifyng existing classifiers in order to curtail the bias induced by imbalanced classes [@estabrooks2004multiple; @chawla2003smoteboost; @chen2004using; @khoshgoftaar2007empirical]. Data level solutions, on the other hand, mostly constitute as re-sampling techniques, in which random sampling is used to either replicate or remove data points in order to create a more balanced dataset [@chawla2004special; @chawla2009data]. It has been shown that quite simple resampling techniques can drastically improve model performance in the face of imbalanced data [@japkowicz2002class; @dubey2014analysis]. For the purposes of this study, we focus our attention toward data level solutions so that all results may be classifier-independent, and, thus, more easily comparable.

Another underexplored aspect of modeling the association between TAD boundaries and genomic annotations is determining a suitable representation of the relationship between them. Typical approaches include overlap counts or percents [@mourad2016computational; @hong2017computational; @huang2015predicting]. Counts refer to the number of times a genomic annotation directly overlaps with a TAD boundary region, e.g., a TAD boundary may overlap with several ChIP-seq peaks for a given transcription factor. Percents refers to the percent of overlap between a genomic annotation and a TAD boundary region, e. g., 60% of a TAD boundary region may overlap with a genomic region annotated as heterochromatin.

While direct overlaps represent the most straightforward measure of functional associations, they do not account for indirect relationships shown to be one of the major driving force of genome regulation [??? REFS]. The most prominent example being the enhancer-promoter relationships, where enhancers are frequently located hundreds or thousands of bases away from their target promoters [@whalen2016enhancer]. On the other hand, the use of distances between TAD boundaries and genomic annotations has largely been ignored. However, these distance features have the potential to offer a more accurate spatial representation of which genomic features are associated with TAD boundaries.

Additionally, model performance is heavily contingent upon the resolution of the contact matrix used to identify the location of TADs throughout the genome. It is often unclear from previous studies what resolution contact matrices were used when retrieving TAD boundaries, and how other resolutions would affect their results. A lower resolution contact matrix will result in the identification of larger-sized TADs, and subsequently, less TAD boundary points. This will contribute to smaller sample sizes and could effect predictive performances.

Therefore, in this study, we aimed to develop an ensemble framework used to build random forest classifiers composed of different combinations of data driven parameters used to address the issues presented above. We used our framework to systematically compare model performances using a composite performance metric, MCC. Our goal was to determine a suitable combination of data re-sampling and feature engineering techniques that both corrected the inherent imbalanced classes and provided the best spatial representation linking genomic annotations to TAD boundaries. Additionally, we used our framework to directly compare the influence of Hi-C contact matrix resolution on model performances. We evaluated Hi-C data for both the GM12878 and K562 cell line obtained at 10 kb, 25 kb, 50 kb, and 100 kb resolution. For each resolution, we examined the inclusion of three different predictor types, including distance as the measure of association between genomic elements. We then tested four different re-sampling techniques aimed at creating balanced classes. Noninformative features were removed using a feature reduction algorithm and random forest classifications were built to assess model performance, and investigate which genomic annotations were most predictive of TAD boundaries for a given system from the complete ensemble.

>>>>

Current methods used to predict TAD boundaries rely on the use of classification algorithms to draw associations between the spatial relationship of functional genomic elements and regions that include TAD boundaries. However, this is a multi-faceted probelem, and many methods ignore key characteristics of domain data when building models and reporting performances. We aim to address these characteristics in order to create a more unified approach toward identifyig the genomic annotations most predictive of TAD boundaries.

A major ignored aspect of TAD boundary prediction is the introduction of heavily imbalanced classes resulting from the number of boundaries vs. nonboundary regions. In a classification setting, when one is attempting to discriminate between regions with (minority class) and without (majority class) TAD boundaries, the class imbalance issue will contribute to small imbalance ratios. The imbalance ratio is defined as the ratio of minority to majority classes [@Orriols-Puig2008]. Such heavily imbalanced datasets are known to adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class; @chen2013influence; @jeni2013facing].

In a study published in 2016 [@mourad2016computational], an L1-regularized multiple logistic regression model was proposed using transcription factor binding sites on GM12878 to discriminate between regions with and without TADs, and to identify the most predictive factors associated with boundaries. However, the class imbalance issue was not addressed. Even worse, when performing cross-validation [@mourad2016computational], used the misclassification error rate of the model to establish the optimal L1-regularization lambda. They also reported ROC (receiver operating characteristic) curves and subsequent AUC (area under the curve) as measures of their model's performance. This is problematic in an imbalanced setting. Such heavily imbalanced datasets are known to adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@lusa2010class; @chen2013influence; @jeni2013facing]. For AUC, when the prevalence of an event is low--that is, a defined genomic region truly containing a TAD boundary--mean probabilities are biased towards the majority class. This results in the area of interest under an ROC curve to be compressed toward a small corner in the ROC space [@prati2011survey; @wei2013role]. While not directly affecting the value of the AUROC, this can have severe consequences on the interpretability of the model's performance. Consequently, careful consideration must be taken when deciding which performance metric(s) to use for evaluation of predictive models, especially in the case of imbalanced data. A better measure to consider is a composite metric involving all quadrants of the confusion matirix, such as Matthew's Correlation Coefficient (MCC) [@jeni2013facing; @chicco2017ten]. The MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications [@brodersen2010balanced]. A value of positive 1 represents a perfect prediction, while a value of negative 1 represents total disagreement between prediction and observation. A value of 0 indicates no better than random prediction.

In addition to employing suitable metrics for evaluating models validated on imbalanced data, there also exists several data level solutions. Data level solutions mostly constitute as re-sampling techniques, in which random sampling is used to either replicate or remove data points in order to create a more balanced dataset [@chawla2004special; @chawla2009data]. It has been shown that quite simple re-sampling techniques can drastically improve model performance in the face of imbalanced data [@japkowicz2002class; @dubey2014analysis].

Two studies published in 2015 and 2017 respectively [@huang2015predicting; @hong2017computational] used re-sampling techniques prior to modeling, and both implemented random under-sampling to balance classes. However, given the large disparity between the number of genomic regions labeled as containing a TAD boundary or not, performing random under-sampling alone is likely to be unstable and introduce bias. Also, it is unclear how other data level solution compare to simple random under-sampling, for instance random over-sampling or a weighted combination of both random under- and over-sampling together. Equally concerning is how the test set is treated in scenarios of re-sampling. It was found that [@huang2015predicting] performed re-sampling prior to splitting the data into training and testing. However, this is problematic because re-sampling prior to splitting can further introduce bias into the testing set, as it is no longer an accurate representation of real-world data [@wei2013role].

Furthermore, determing the optimal representation of the relationship between features in the model and genomic regions, known as feature engineering, plays a significant role in TAD prediction. Typical approaches include overlap counts [@hong2017computational] or percents [@mourad2016computational; @huang2015predicting]. Counts refer to the number of times a genomic annotation directly overlaps with a TAD boundary region, e.g., a TAD boundary may overlap with several ChIP-seq peaks for a given transcription factor. Percents refers to the percent of overlap between a genomic annotation and a TAD boundary region, e. g., 60% of a TAD boundary region may overlap with a genomic region annotated as heterochromatin. While direct overlaps represent the most straightforward measure of functional associations, they do not account for indirect relationships shown to be one of the major driving force of genome regulation [??? REFS]. The most prominent example being the enhancer-promoter relationships, where enhancers are frequently located hundreds or thousands of bases away from their target promoters [@whalen2016enhancer]. Using percents offers a more granular approach, but it is unclear how it performs when compared to counts. On the other hand, the use of distances between TAD boundaries and genomic annotations has largely been ignored. However, these distance features have the potential to offer the most accurate spatial representation of which genomic features are associated with TAD boundaries.

Additionally, model performance is heavily contingent upon the resolution of the contact matrix used to identify the location of TADs throughout the genome. It is often unclear from previous studies what resolution contact matrices were used when retrieving TAD boundaries, and how other resolutions would affect their results. A lower resolution contact matrix will result in the identification of larger-sized TADs, and subsequently, less TAD boundary points. This will contribute to smaller sample sizes and could effect predictive performances. To date, no study has compared the results of their predictive models built across multiple data resolutions.

Therefore, in this study, we aimed to develop an ensemble framework composed of different combinations of data driven characteristics used to address the issues presented above. We used our framework to systematically compare performances of random forest classifiers using a composite performance metric, MCC. Our goal was to determine a suitable combination of data re-sampling and feature engineering techniques that both corrected the inherent imbalanced classes and provided the best spatial representation linking genomic annotations to TAD boundaries. Additionally, we used our framework to directly compare the influence of Hi-C contact matrix resolution on model performances. We evaluated Hi-C data for both the GM12878 and K562 cell line obtained at 10 kb, 25 kb, 50 kb, and 100 kb resolution. For each resolution, we examined the inclusion of three different predictor types, including distance as the measure of association between genomic elements. We then tested four different re-sampling techniques aimed at creating balanced classes. Noninformative features were removed using a feature reduction algorithm and random forest classifications were built to assess model performance, and investigate which genomic annotations were most predictive of TAD boundaries for a given system from the complete ensemble.
